Preregistration for modelling in ecolgy and conservation

### 2023-05-15 EcoConsPreReg-supervisor_draft-Sep-2020.Rmd
update manuscript introduction

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elise Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
bibliography: ./data/references.bib
output: word_document
  #distill::distill_article:
    #self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

----

## Introduction ##

### Problem ###

Ecology as a discipline is at risk of a "reproducibility crisis" (Fidler et al. 2017), with rates of Questionable Research Practices (QRPs) in ecology similar to those in other fields (Fraser et al. 2018), such as psychology and medicine, where the crisis has received considerable attention, and provoked heated discussion among researchers within the broader scientific community. In particular, there are serious problems with the way ecological models and the modelling process are reported in the published literature. However, concerns about poor model reporting practices are not new, and have been raised repeatedly --- nearly a decade ago researchers concluded that most model reporting was neither transparent or comprehensive [@Schmolke:2010fd; @deVos:2011], and that a culture of model reporting was yet to be established [@Grimm:2014es]. Moreover, most models are assessed predominantly on the basis of scientific originality during peer-review  [@Schmolke:2010fd] to the effect that the burden of evaluating models for credibility is placed onto the reader or model-user, rather than the peer-review process, especially in the case of extremely complex models [@Alexandrov:2011iv].

Despite attempts to promote thorough and transparent model documentation using 'Good Modelling Practice' (e.g. [@Jakeman:2006ii;@Alexandrov:2011iv]), more recent quantitative assessments of transparency and reproducibility demonstrate that transparent model reporting is an ongoing problem. Just as Jakeman argued that many models in the published literature lack a clear statement of model objectives and purpose, lack descriptions of model assumptions and implications, and also fail to report model results, including validation and evaluation in 2006, [-@Jakeman:2006ii], [@Feng2019]'s quantitative assessment of Ecological Niche Models found that reporting of model evaluation and model results was less than adequate.

A lack of complete and transparent reporting practices is concerning because it  provides the conditions in which QRPs are likely to arise [@Fidler:2017he]. Moreover, as Morrison [-@Morrison:2016cd] demonstrated, when models are not reported transparently with sufficient detail, they cannot be reproduced. This is particularly problematic because in ecology, models are typically used in decision-support, often directly underpinning policy and management decisions. Incomplete and non-transparent reporting of the model and the modelling process risk the model being used in inappropriate applications or decision contexts [@Grimm:2014es] because the underlying modelling choices and assumptions are not communicated [@deVos:2011]. Alternatively, a lack of transparency may mask poor model design or serious flaws in the model, with the risk that they be used in decision support resulting in wrong, or irreversible decisions [@Grimm:2014es;@Schmolke:2010fd]. A lack of transparency in reporting also risks undermining the credibility of the model, such that the decision-maker rejects the use of the model and instead turns to unstructured and subjective judgments [@Grimm:2014es;@Addison:2013wa], thus contributing further to "decision complacency", or the "knowing-doing" gap [@Gardner:2018dm].

Preregistration is a practice that has been adopted in disciplines that have confronted the 'reproducibility crisis' head-on, and can improve the transparency and reproducibility of ecological models. Preregistration requires the methods and analysis plan of a future study are registered in a secure and accessible platform prior to any data collection, and after submission it can no longer be altered. Preregistration distinguishes between what is a genuine a priori planned analysis, and what is not (Parker, Fraser, and Nakagawa 2019), and therefore prevents the opportunistic misuse of researcher degrees of freedom or 'Questionable Research Practices' (QRPs, Wicherts et al., 2016). Researcher degrees of freedom refer to the many analytic decisions a researcher makes during the planning, data collection, data analysis and reporting of results in scientific studies (Wicherts et al. 2016; Simmons, Nelson, and Simonsohn 2011). Examples of these analytic choices include whether to exclude observations, whether more data should be collected, which set of conditions should be combined and compared, which model evaluation metric should be used, etc. Coupled with a lack of transparency -- or "undisclosed flexibility" -- researcher degrees of freedom are vulnerable to being exploited by researchers because of cognitive biases and incentive structures within science. \<lead in sentence to next para??\>

Ecology has seen recent improvements in journal policy concerning reporting guidelines and data/code archiving, but there has been slow uptake of other important tools, like preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines to think that preregistration is "not for them", this was a particularly salient sentiment for researchers engaged in non-experimental studies. We have experienced this sentiment ourselves when looking at preregistration templates to use for model-based research --- nearly all preregistration templates are geared towards experimental hypothesis-testing research. Such templates contain items that are simply irrelevant to model-based and/or applied ecological research. More importantly, they do not reflect the iterative nature of model development [@MacEachern2019] and fail to capture critical analytic decisions in the modelling process. Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however it is steeped heavily in an an academic psychology context, is focussed on the use of models for hypothesis-testing, neglects the8 model development process, and approaches modelling from a very narrow epistemic function, in a way that is not commonly applied in ecology.

Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology. Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology.

### Research Aims, and Template Objectives ###

The goal of this study is to \<rethink\> preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology. This research builds on recently developed "QRP roadmaps" (Gould et al. 2019) and research by Cruwell et al. (2019) showing that existing templates are too general, and do not account for domain-specific and research category-specific nuances and questionable research practices. We aim to develop a template for preregistering ecological models, and devise a methodology for implementing an expanded view of the preregistration process, 'adaptive preregistration' (*sensu* Srivastava), that better accounts for the inherently iterative nature of model development than static, deterministic document-based preregistration.

We aim to consider questions such as... (open questions we had about preregistration)...

Finally, we develop and evaluate both the template and proposed adaptive preregistration process with a case-study of environmental-flow modelling in Victoria, Australia.

----

## Methods ##

### Designing Preregistration Templates with a User-Centred Design Approach ###

@@

We define the primary purpose of the template as "to delimit researcher degrees of freedom" when conducting model-based research in applied ecology and conservation.  Although the purpose of preregistration is predominantly conceptualised in terms of distinguishing between exploratory from confirmatory research [ e.g. @Nosek:2018eg], we wish to emphasise that this is not the intended purpose of this template.  We adopt a more general stance towards the function of preregistration, i.e. that preregistration should mitigate against questionable research practices by reducing the potential for cognitive biases to cause opportunistic misuse of researcher degrees of freedom, such as cherry-picking and HARKing.

@@

c) Use this knowledge to inform the design of the templates so they meet their intended purpose and are taken up by researchers.

@@

#### Analysis and Design ####

@@

----

## Results: A methodology for Preregistration in Ecology and Conservation ##

----

### Internal Model of Preregistration ###

@@

In order to address the disjuncture between the iterative nature of model development, and the single-use, deterministic preregistration template, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by \<Srivastava\>. This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' \<Srivastava\> wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed \<Srivastava\>. Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed \<Link to Figure, GitHub as a platform for adaptive preregistration, <x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0>\>.

@@

In redesigning the procedure and interface for completing a preregistration using an adaptive preregistration process, we followed the advice by *Pu et al. [-@Pu2019]* who recommended two design elements that may facilitate checking. Firstly, that the design should "make relevant preregistration easier to *query"* and secondly it should "encourage wider *coverage* of preregistration content during review." The process of *querying* a preregistration consists of "paper-to-preregistration" and "preregistration-to-paper" lookup whereby the reader searches for relevant information from the paper in the preregistration and *vice-versa*. This requires that the report or paper must be specific in referring to parts of the preregistration, *and*, that the author has tagged the relevant part of the preregistration in their report or manuscript.

@@

----

### Adaptive Preregistration for model-based research using GitHub ###

@@

GitHub is used to track amendments to the preregistration --- each time an additional question is answered, or an alteration is made to the preregistration, the change should be *committed* to the *repository*. *Semantic versioning* is used to document each version of the preregistration.  *Semantic versioning* is an open-source software engineering practice used to document and and communicate the development of software, in a way that allows others to depend on it [@Kitzes2018vn;@Raemaekers2017]. We have created our own *semantic versioning* scheme to be used to specifically document the incremental version of the preregistration, in tandem with *GitHub's* tag and release feature, each interim preregistration is tagged and released with its semantic version number.

@@

----

### A Shiny Preregistration Interface ###

@@

----

## Case Study ##

@@

----

## Discussion ##

### talking points ###

@@

**[@MacEachern2019]:**

**"**We fear that preregistration will lessen the engagement of analysts with data and that it represents a step back toward the punch card style of analysis." MacEachern, S. N., & Van Zandt, T. (2019). Preregistration of Modeling Exercises May Not Be Useful. Comput Brain Behav, 2(3-4), 179-182. doi:10.1007/s42113-019-00038-x

@@

Moreover, our design template is *opinionated*. (See Elith's latest SDM reporting guideline) -\> it provides a *guide*. For those whom are at risk of 'punch card style analysis', ala MacEachern.  
--------------------------
# References

>>>>

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elliot Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
abstract: "Blah blah"
date: "`r Sys.Date()`"
bibliography: ./data/references.bib
output: #
  distill::distill_article:
  self_contained: false
---

**Intended journal:** Methods in Ecology and Evolution.

@@

------------------------------------------------------------------------

## Introduction

Preregistration is a practice that has been adopted in disciplines that have confronted the 'reproducibility crisis' head-on, and aims to improve the transparency and reproducibility of ecological models. Preregistration requires the methods and analysis plan of a future study are registered in a secure and accessible platform prior to any data collection, and after submission it can no longer be altered. Preregistration distinguishes between what is a genuine a priori planned analysis, and what is not (Parker, Fraser, and Nakagawa 2019), and therefore prevents the opportunistic misuse of researcher degrees of freedom or 'Questionable Research Practices' (QRPs, Wicherts et al., 2016). Researcher degrees of freedom refer to the many analytic decisions a researcher makes during the planning, data collection, data analysis and reporting of results in scientific studies (Wicherts et al. 2016; Simmons, Nelson, and Simonsohn 2011). Examples of these analytic choices include whether to exclude observations, whether more data should be collected, which set of conditions should be combined and compared, which model evaluation metric should be used, etc. Coupled with a lack of transparency -- or "undisclosed flexibility" (cite)-- researcher degrees of freedom are vulnerable to being exploited by researchers because of researchers' cognitive biases and incentive structures within science.

Ecology has seen recent improvements in journal policy encouraging or mandating open-scientific reforms that aim to increase transparency and reproducibility of published research, for instance, reporting guidelines and data/code archiving. However, there has been slow uptake of other important open-science tools, such as preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines adopting preregistration in their scientific practice, because they consider preregistration is "not for them". We have experienced this sentiment ourselves when considering preregistering model-based research because nearly all preregistration templates are focussed on experimental hypothesis-testing research, and as such do not reflect the research life-cycle of model-based research; they fail to capture critical decisions commonly implemented in the model-development process, or include fields that relate to analytic decisions that are simply irrelevant to model-based research. Moreover,

Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however the template explicitly excludes the construction and development phases of modelling, because the authors adopt a narrow but dominant view pervading the published literature on preregistration, 1. that scientific inquiry can be divided wholly into confirmatory and exploratory research, and that 2. that preregistration is intended to delimit researcher degrees of freedom within hypothesis-testing research only. The exclusion of preregistration from model-based research is premised on the idea, summarised by MacEachern and Van Zandt [-@MacEachern2019], that the modelling process can be dichotomised into two distinct phases that map onto the exploratory/confirmatory research divide, wherein model-development is inherently exploratory, 'creative', and 'confirmatory modelling' occurs within the context of experimental studies and the model is treated as a hypothesis to be tested when confronted with new data.

In this paper, we take a radical departure from this perspective of preregistration, arguing that preregistration can and should be used for model-based research, including model development and construction, however, preregistration should reflect the norms and practices of the research context in which it is applied, in order for it to be broadly accepted and adopted by any given community of practice. Dichotomising model-based research into either exploratory or confirmatory research is reductive, and does not reflect the domain-specific uses of modelling in ecology and conservation in particular (For example, see [@Brudvig2017] for one epistimology describing the uses of modelling in applied and basic restoration) (**TODO elaborate further??**).

Crucially, we believe it's problematic to preclude the model-development and construction phases of modelling from preregistration, because questionable research practices (QRPs) are possible (and likely just as prevalent for other types of ecological and scientific research), and preregistration is a valuable tool in mitigating the risk of QRPs in model-based research. Thirdly, preregistration provides additional benefits, and can improve the transparency of decision-making during modelling, which is notoriously opaque and often under-described (**TODO insert citation / refs on the lack of transparency in model based research).**

The goal of this study is to rethink preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology and conservation. This research builds on recently developed "QRP roadmaps" (Gould et al. 2019) and research by Cruwell et al. (2019) showing that existing templates are overly-focussed on hypothesis-testing research, and do not account for domain-specific and research category-specific nuances and questionable research practices. Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology.

We first aim to devise a methodology for implementing an expanding view of the preregistration process, 'adaptive preregistration' [sensu @Srivastava2018], that better accounts for the inherently iterative nature of model development than the prevailing immutable, unalterable and deterministic view of preregistration. Adaptive preregistration can be applied to model-based research because it allows for the iterative cycle of model development to be captured in the preregistration.

Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology. Secondly, we aim to develop a preregistration template for ecological modelling. Finally, we evaluate both the template and proposed procedure for adaptive preregistration using a case-study in which we preregister and undertake an environmental flows modelling study from Victoria, Australia.

## Methods

### Designing Preregistration Templates with a User-Centred Design Approach

@@

We define the primary purpose of the template as "to delimit researcher degrees of freedom" when conducting model-based research in applied ecology and conservation. Although the purpose of preregistration is predominantly conceptualised in terms of distinguishing between exploratory from confirmatory research [e.g. @Nosek:2018eg], we wish to emphasise that this is not the intended purpose of this template. We adopt a more general stance towards the function of preregistration, i.e. that preregistration should mitigate against questionable research practices by reducing the potential for cognitive biases to cause opportunistic misuse of researcher degrees of freedom, such as cherry-picking and HARKing.

@@

d)  Use this knowledge to inform the design of the templates so they meet their intended purpose and are taken up by researchers.

@@

#### Analysis and Design

@@

------------------------------------------------------------------------

## Results: A methodology for Preregistration in Ecology and Conservation

------------------------------------------------------------------------

### Internal Model of Preregistration

@@

In order to address the disjuncture between the iterative nature of model development, and the single-use, deterministic preregistration template, we take an 'expanded view of preregistration', that fits with the concept of 'adaptive preregistration' proposed by \<Srivastava\>. This view of preregistration breaks with the current model of preregistration, wherein the author writes a single deterministic preregistration containing a rule for every decision, and shifts from ideation, preregistration to execution of the plan. Two features define adaptive preregistration. It contains 'plans to deploy flexible strategies' \<Srivastava\> wherein the author can supply a heuristic consisting of multiple different analysis or modelling strategies whose execution depends on the outcome of previous decision-points or analyses. Secondly, the preregistration may be iterative and consist of interim preregistrations that mark phases of modelling and analysis as different parts of the data are observed \<Srivastava\>. Thus, as the modeller proceeds through the model development process, the researcher can shift from ideation, preregistration, execution of the analysis and back to ideation and preregistration again, depending on the observed outcomes as each interim preregistration is executed \<Link to Figure, GitHub as a platform for adaptive preregistration, \<x-scrivener-item:///Users/egould/Dropbox/Apps/Scrivener/PhD_thesis.scriv?id=32FCFBDA-7CEE-4FD9-B8DD-F3073309C0D0\>\>.

@@

In redesigning the procedure and interface for completing a preregistration using an adaptive preregistration process, we followed the advice by *Pu et al. [-@Pu2019]* who recommended two design elements that may facilitate checking. Firstly, that the design should "make relevant preregistration easier to *query"* and secondly it should"encourage wider *coverage* of preregistration content during review." The process of *querying* a preregistration consists of "paper-to-preregistration" and "preregistration-to-paper" lookup whereby the reader searches for relevant information from the paper in the preregistration and *vice-versa*. This requires that the report or paper must be specific in referring to parts of the preregistration, *and*, that the author has tagged the relevant part of the preregistration in their report or manuscript.

@@

------------------------------------------------------------------------

### Adaptive Preregistration for model-based research using GitHub

@@

GitHub is used to track amendments to the preregistration --- each time an additional question is answered, or an alteration is made to the preregistration, the change should be *committed* to the *repository*. *Semantic versioning* is used to document each version of the preregistration. *Semantic versioning* is an open-source software engineering practice used to document and and communicate the development of software, in a way that allows others to depend on it [@Kitzes2018vn; @Raemaekers2017]. We have created our own *semantic versioning* scheme to be used to specifically document the incremental version of the preregistration, in tandem with *GitHub's* tag and release feature, each interim preregistration is tagged and released with its semantic version number.

@@

------------------------------------------------------------------------

### A Shiny Preregistration Interface

@@

------------------------------------------------------------------------

## Case Study

@@

------------------------------------------------------------------------

## Discussion

### talking points

@@

[@MacEachern2019]**:**

**"**We fear that preregistration will lessen the engagement of analysts with data and that it represents a step back toward the punch card style of analysis." MacEachern, S. N., & Van Zandt, T. (2019). Preregistration of Modeling Exercises May Not Be Useful. Comput Brain Behav, 2(3-4), 179-182. <doi:10.1007/s42113-019-00038-x>

@@

## Moreover, our design template is *opinionated*. (See Elith's latest SDM reporting guideline) -\> it provides a *guide*. For those whom are at risk of 'punch card style analysis', ala MacEachern.

# References
